\documentclass[]{llncs}
\title{A Formal Connection between Security Automata and JML
Annotations\thanks{This work is partially funded by the IST FET
programme of the European Commission, under the IST-2005-015905
\textsf{Mobius} project. Research done while the authors where at INRIA Sophia
Antipolis.}}

\author{Marieke Huisman\inst{1}
 \and Alejandro Tamalet\inst{2}}
% \author{Marieke Huisman\inst{1}\thanks{Research done while at INRIA Sophia Antipolis}
%  \and Alejandro Tamalet\inst{2}\thanks{Research done while at INRIA Sophia Antipolis}}

\institute{University of Twente, Netherlands \and
University of Nijmegen, Netherlands}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{psfrag}
\pagestyle{plain}
\usepackage{xspace}
\input{commands}
\usepackage{paralist}
\usepackage{url}
\usepackage{floatflt}
\usepackage{longtable}
\usepackage{alltt}
\begin{document}

\maketitle
\begin{abstract}
Security automata are a convenient way to describe security
policies. Their typical use is to monitor the execution of an
application, and to interrupt it as soon as the security policy is
violated. However, run-time adherence checking is not always
convenient. Instead, we aim at developing a technique to verify
adherence to a security policy statically.  To do this, we consider
a security automaton as specification, and we generate JML
annotations that inline the monitor~--~as a specification~--~into the
application.  We describe this translation and prove preservation of
program behaviour, \emph{i.e.}, if monitoring does not reveal a
security violation, the generated annotations are respected by the
program.

The correctness proofs are formalised using the PVS theorem prover.
This reveals several subtleties to be considered in the definition of the
translation algorithm and in the program requirements.
\end{abstract}


\input{introduction}
\input{mva}
\input{program}
\input{annotgen}
\input{relatedwork}
\input{concl}

\bibliographystyle{plain}
\bibliography{specification}

% \vspace*{-0.7em}
\subsubsection*{Acknowledgements}
We thank Erik Poll for his useful comments on an earlier draft of this
paper, and Igor Siveroni, who started the work on this topic and came
up with the idea to use method-level set-annotations.
\end{document}
