\documentclass{llncs}
\usepackage{times}
\usepackage{epsfig}
\usepackage{enumerate}
\usepackage{paralist}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amscd}
\usepackage{array}
\usepackage{tabularx}
\usepackage{xspace}
\usepackage{url}
\usepackage{float}
\usepackage{multirow}
\usepackage{listings}
\usepackage{latexsym,pstricks,pst-node,pst-tree,boxedminipage,stmaryrd}
\usepackage{alltt}
\usepackage{fancyhdr}
%\usepackage{fullpage}

%\markboth{INSPIRED}{INSPIRED}

%\pagestyle{myheadings}

\pagestyle{plain}
\title{Methodology for Java Application Validation}

\author{}

\input {preamble}          % input user-defined commands

\begin{document}
\maketitle      
\begin{abstract}
This article presents a developer
methodology for developer oriented application validation based on 
the Jack validation framework. The tool allows to prove properties on
Java program in a user-friendly environment. Java program annotated
with JML assertions can be proved correct at source or at bytecode
level. Many phases of the validation are automated: annotation
generation, anotation propogation, verification condition generation,
proof. The proof part is and cannot be fully automated and different
interactive provers are linked to the tool allowing to prove remaining
lemmas. The tool is fully integrated in the eclipse IDE. A Coq proof
environment has also been developped as an eclipse plugin in order to
have a unique interface in all the validation phases.

Benefiting from users feedback, the tools have also been improved
to deal with different purposes such that security properties
verification, functional properties verification, code optimization,
or resource consumption verification.
\end{abstract}
\input{introduction}        
\input{methodology}        
\input{evaluation}
\input{conclusion}

\bibliographystyle{plain}
\bibliography{biblio,bib_mem-wpc,sttt,specification,bibliography,jack,../specification.bib,/net/home/gbarthe/bib/string,/net/home/gbarthe/bib/gilles,/net/home/gbarthe/bib/article,/net/home/gbarthe/bib/book,/net/home/gbarthe/bib/lncs,/net/home/gbarthe/bib/misc,/net/home/gbarthe/bib/proceedings,/net/home/gbarthe/bib/techrep,/net/home/gbarthe/bib/thesis,/net/home/gbarthe/bib/software,/net/home/gbarthe/bib/web}

\end{document}

