
%\section{Preliminaries} 
% \subsection{JACK}
%Jack \footnote{ http://www-sop.inria.fr/everest/soft/Jack/jack.html} is a plugin for the eclipse \todo{referenced - the eclipse site} integrated development environment for Java. Originally, Jack statically checks, using formal methods, source programs against their JML specification. The tool can interface with several theorem provers (AtelierB, 
%Simplify, Coq, PVS ) \todo{references}. We have upgraded the tool with implementations of the JML compiler and the bytecode verification condition generator. Jack has a user-friendly interface which makes it easy for use by developers that do not have a large background in formal methods.


\section{Overview of JML} \label{BCSLprelim}
% what is JML
JML~\cite{JMLRefMan} (short for Java Modeling Language) is a behavioral interface specification 
language tailored to Java applications which follows the design-by-contract approach (see~\cite{M97oos}).

% useful language
Over the last few years, JML has become the de facto specification language for
 Java source code programs. Several case studies have demonstrated that JML can be used to specify realistic
industrial examples, and that the different tools allow to find errors
in the implementations (see~e.g. \/~\cite{BreunesseCHJ04}). One
of the reasons for its success is that JML uses a Java-like
syntax.  Other important factors for the success of JML are its expressiveness and
flexibility.

% tool support 
JML is supported by several verification tools.
 Originally, it has been designed as a language of the runtime assertion checker~\cite{jmlrac} created by G.T. Leavens and \todo{who are the others}. 
The JML rac compiles both the Java code and the JML specification into executable bytecode and thus, 
in this case, the verification consists in executing the resulting bytecode. Several static checkers based 
on formal logic exist which use JML as a specification language. Esc/java~\cite{escjava}  whose first version used a subset 
 of JML \footnote{the current version of the tool esc/java 2  supports almost all JML constructs} is among the first tools supporting JML.
Among the static checkers with JML  are
the Loop tool developed by the Formal group at the University of Nijmegen,
the Jack tool developed at Gemplus, the Krakatoa tool created by the Coq group at Inria France.
The tool Daikon \cite{ECG01DDL}  tool uses a subset of JML for detecting loop invariants by run of programs.
 A  detailed overview of the tools which support JML can  be found in~\cite{BurdyCCEKLLP03}.



% technical details
    Specifications are written using different predicates
    which are side-effect free Java expressions, extended with
    specification-specific keywords. JML specifications are written as comments so they are not
    visible by Java compilers. The JML syntax is close to the Java syntax: JML extends the Java
    syntax with few keywords and operators. 
    For introducing method precondition and postcondition the keywords \jmlKey{requires} and \jmlKey{ensures} are used  
    respectively,  \jmlKey{modifies} keyword is followed by all the locations that can be modified by the method, 
    \jmlKey{loop\_invariant}, stands for loop invariants, the \jmlKey{loop\_modifies} keyword gives the 
    locations modified by loop invariants etc. The latter is not standard in JML and is an extension introduced in~\cite{BRL-JACK}.
    Special JML operators are, for instance, \jmlKey{$\backslash$result} which stands for the value that a method returns if it is not
    void, the \jmlKey{$\backslash$old(expression)} operator 
    designates the value of \jmlKey{expression} in the prestate of a method and is usually
    used in the method's postcondition. JML also allows the declaration of special JML variables, that are used only for specification purposes. 
    These variables are declared in comments with the \jmlKey{model} modificator and may be used only in specification clauses. 



     
Figure~\ref{replaceSrc} gives an example of a Java class that models a list stored in a private array field. 
The method \texttt{replace} will search in the array for the first occurence of the object \texttt{obj1} passed as first argument and if found, it will be replaced with the object passed as second argument \texttt{obj2} and the method will return true; otherwise it returns false. The loop in the method body has an invariant which states that all the elements of the list that are inspected up to now are different from the parameter object \texttt{obj1}. The loop specification also states
that the local variable \texttt{i} and any element of the array field \texttt{list} may be modified in the loop.

\begin{figure}[ht!]
\begin{lstlisting}[frame=trbl] 

public class ListArray {

  private Object[] list;
  
  //@ requires list != null;
  //@ ensures \result ==(\exists int i; 
  //@ 0 <= i && i < list.length && 
  //@ \old(list[i]) == obj1 && list[i] == obj2);
  public boolean replace(Object obj1,Object obj2){
    int i = 0;
    
    //@ loop_modifies i, list[*];
    //@ loop_invariant i <= list.length && i >=0 
    //@ && (\forall int k;0 <= k && k < i ==> 
    //@     list[k] != obj1);
    for (i = 0; i < list.length; i++ ){
      if ( list[i] == obj1){
        list[i] = obj2;
        return true;	
      }
    }
    return false;
  }
}
\end{lstlisting}
\caption{\sc class \mbox{\rm \lstinline!ListArray!} with JML annotations} 
\label{replaceSrc}
\end{figure}



% light and heavy weight specification
A useful feature of JML is that it allows two kinds of method specification, a \light  \ and \heavy \ weight specification. 
An example for a \light \  specification is the annotation of method \texttt{replace} in Fig. \ref{replaceSrc}. The specification in 
the example states what is the expected behavior of the method and under what desired conditions it might be called.
The user, however, has also the possibility
to write very detailed method specifications in JML. This style of specification is called a heavy weight specification.
 It is introduced by the JML keywords \jmlKey{normal\_behavior} and \jmlKey{exceptional\_behavior}. As the keywords
suggest every of them specifies a specific normal or exceptional behavior of a method.  (see \cite{PD06LBR}). 
The keyword  \jmlKey{normal\_behavior} introduces a precondition which guarantees that if it holds in the prestate of the
 method execution then the method will terminate normally and the postcondition introduced by the keyword \jmlKey{ensures} will hold.
 Note that this clause guarantees that the method will not terminate on an exception and thus the exceptional postcondition 
for any kind of exception (i.e. for the exception class  \texttt{Exception}) is  \jmlKey{false}.
% example for heavy weight specification
An example for a \heavy \ weight specification is given in Fig. \ref{bml:heavySp}. In the example, method \texttt{divide} has 
two behaviors, one in case the method terminates normally and the other in case the method terminates by throwing an object reference of
 \texttt{ArithmeticException}. The exceptional postcondition is ommitted in the normal behavior specification clause as by default
if the precondition \texttt{ b > 0} holds this assures that no exceptional termination is possible. 
Another observation over the example is that the exceptional behavior is introduced with the JML keyword \jmlKey{also}. 
The keyword  \texttt{also} serves for introducing every new behavior of a method. Note that the keyword \jmlKey{also} is used in case a  method overrides
a method from the super class.
 In this case, the method specification is preceded by the keyword \jmlKey{also} to indicate that the method should respect also the specification 
of the super method.


   
\begin{figure}
\begin{lstlisting}[frame=trbl]
public class C {
    int a;
    
    //@ public instance invariant a > 0 ;
    
    //@ requires val > 0 ;
    public C(int val){
       a = val ;
    }
   
    //@ public normal_behavior
    //@ requires b > 0;
    //@ modifies a;
    //@ ensures  a == \old(a) / b;  
    //@
    //@ also 
    //@ public exceptional_behavior
    //@ requires b == 0;
    //@ modifies \nothing;
    //@ exsures (ArithmeticException) a == \old(a);
    public void divide(int b) {
        a = a / b;
    }
}
\end{lstlisting}
\caption{\sc An example for a method with a heavy weight specification in JML} \label{bml:heavySp}
\end{figure}

JML can be used to specify not only methods but also properties of a class or interfaces. 
A Java class may be specified with an invariant or history constraints. An invariant of a class is a predicate which holds at all visible states of
  every object of this class (see for the definition of visible state in the JML reference manual \cite{december-jml}). 
  A Class history constraints is a property which relates the initial and terminal state of every method in the corresponding class. 
 The class \texttt{C}  in Fig.\ref{bml:heavySp} has also an invariant which states that the instance variable \texttt{a} is
 always greater than \texttt{0}.


