\documentclass[]{llncs}

\title{FMCO Tutorial Proposal:\\
JACK~---~a tool for validation of security and behaviour of Java applications}

\author{Gilles Barthe\inst{1} \and
        Lilian Burdy\inst{1} \and
        Julien Charles\inst{1} \and
        Benjamin Gr\'egoire\inst{1} \and
        Marieke Huisman\inst{1} \and
        Jean-Louis Lanet\inst{2} \and
        Mariela Pavlova\inst{1} \and
        Antoine Requet\inst{2}}
\institute{INRIA Sophia Antipolis, France \and Gemplus, France}

\begin{document}
\maketitle

%\begin{abstract}
%JACK is a tool set for the validation of Java applications. The
%security requirements and behaviour of the application can be
%specified with JML. JACK computes the verification conditions that
%ensure correctness of the application. The verification conditions can
%be verified with different automatic and interactive provers.

%To support the whole development process of secure applications, JACK
%allows to do the verification both at source code and at bytecode
%level. Specifications of source code programs can be automatically
%compiled into bytecode specifications. 

%JACK also provides support for writing specifications, as it
%implements several algorithms to automatically generate annotations.
%\end{abstract}

\paragraph{Motivation}
Over the last years, the use of trusted personal devices, such as
mobile phones, PDAs and smart cards, has become more and more
widespread. As they are often used with security-sensitive
applications, they are an ideal target for attacks. Traditionally,
research has been focusing on avoiding hardware attacks, but the
emergence of new generation trusted personal devices that are
increasingly connected to networks and provide execution support for
complex programs has put forward the need to improve the quality of
software to avoid logical attacks.  Logical attacks are potentially
easier to launch than physical attacks (for example they do not
require physical access to the device, and are easier to replicate
from one device to the other), and may have a huge impact.  In
particular, a malicious attacker spreading over the network and
disconnecting or disrupting devices massively could have deep
consequences.

\paragraph{A tool for application validation}
We report here on the development of the tool JACK that can be used
throughout the application development process to increase confidence
in it. JACK allows to verify Java applications that are annotated with
JML (Java Modeling Language\footnote{See
\texttt{http://www.jmlspecs.org}.}). An advantage of using JML is that
there is wide range of tools and techniques available that use JML as
input language, \emph{i.e.}\ JML specifications can be used to improve
testing, they can be simulated, and in particular they can be
verified. We distinguish two kinds of verification: at runtime, using
\texttt{jmlc} or statically. Several tools provide
static verification of JML-annotated programs, adopting different
compromises between soundness and automation.  Among these tools, JACK
aims at achieving a high level of soundness, while keeping
automation. Therefore, JACK automatically generates proof obligations
that can be discharged both by automatic and by interactive theorem
provers. The automatic prover that is used is Simplify, that is also
used by several other tools for program verification. The interactive
theorem prover that is used by JACK is Coq. 

The tutorial will focus on the main characteristics of JACK that
distinguish it from other tools for the static verification of
JML-annotated Java programs is that it provides support for the whole
development process of secure applications:
\begin{itemize}
\item it is integrated with the widely-used IDE Eclipse\footnote{See
\texttt{http://www.eclipse.org}.};
\item it provides means to generate annotations; and
\item it allows verification of source code and bytecode level
programs, and source code specifications can be compiled into bytecode
specifications.
\end{itemize}
Moreover, JACK allows to verify complex functional behaviour
specifications, by providing advanced support for interactive
verification. 


\paragraph{Integration with Eclipse}
Because of the tight integration with Eclipse, the developer does not
have to change tools to validate the application. A special JACK view
has been implemented, that allow to inspect the generated proof
obligations in different views (in a Java-like syntax, or in the
language of the prover). Moreover, syntax colouring of the original
source code allows to see to which parts of the application and
specification the proof obligation relates.

\paragraph{Generation of annotations}
One of the drawbacks of using JML-like annotations for specification
is that it is labour-intensive and error-prone to write them, as it is
easy to forget some annotations. We provide two different annotation
generators: the first generator computes frame conditions and simple
preconditions that are sufficient to avoid runtime exceptions, the
second generator takes as input a security policy and generates and
propagates annotations in such a way that if the application respects
the annotations, then it also respects the security policy.

\paragraph{Verification of bytecode programs}
A possible approach to the secure loading of applications on trusted
personal devices is the use of proof-carrying code. In such a
framework, the applications come equipped with a specification and a
proof that allow the client to establish trust in the application. To
achieve this, we need the ability to reason about bytecode. JACK
provides a verification condition generator for BML, the Bytecode
Modeling Language, a bytecode variation of BML. In addition, it
provides a compiler from JML to BML, and we have shown that this
compiler basically preserves proof obligations (and thus source code
level proofs can be reused at bytecode level).

\paragraph{Support for interactive verification}
As interactive program verification is known to be labour-intensive,
several advanced Coq tactics have been developed that help with this
verification. To be able to write expressive specifications, a
\texttt{native} construct has been proposed for JML, that allows to
link JML constructs directly with the logic of the underlying
prover. This allows to develop the theory about these constructs
directly in the logic of the theorem prover, which makes verification
much simpler.

\paragraph{Literature}
Several papers have been published around JACK; here we list a few.

\nocite{BRL03:fme,Charles06,m+04:cardis,BurdyP06}

\bibliographystyle{plain}
\bibliography{../specification.bib,/net/home/gbarthe/bib/string,/net/home/gbarthe/bib/gilles,/net/home/gbarthe/bib/article,/net/home/gbarthe/bib/book,/net/home/gbarthe/bib/lncs,/net/home/gbarthe/bib/misc,/net/home/gbarthe/bib/proceedings,/net/home/gbarthe/bib/techrep,/net/home/gbarthe/bib/thesis,/net/home/gbarthe/bib/software,/net/home/gbarthe/bib/web}

\end{document}
