
\documentclass{llncs}
%\usepackage{times}
\usepackage{epsfig}
%\usepackage{enumerate}
%\usepackage{paralist}
%\usepackage{amsmath}
%\usepackage{amssymb}
%\usepackage{amscd}
%\usepackage{array}
%\usepackage{tabularx}
\usepackage{xspace}
\usepackage{url}
%\usepackage{float}
\usepackage{multirow}
\usepackage{listings}
%\usepackage{latexsym,pstricks,pst-node,pst-tree,boxedminipage,stmaryrd}
%\usepackage{alltt}
%\usepackage{fancyhdr}
%\usepackage{fullpage}

\usepackage{paralist}
%\markboth{INSPIRED}{INSPIRED}

%\pagestyle{myheadings}

% julien: mes macros
\newcommand{\rarrow}{$\rightarrow$}
\newcommand{\conj}{$\wedge$}
\newcommand{\disjonc}{$\vee$}
\newcommand{\s}{\,}
\newcommand{\btab}{\begin{tt}\begin{tabbing}}
\newcommand{\etab}{\end{tabbing}\end{tt}}
% julien: fin de mes macros
\newcommand{\native}{\texttt{native}\xspace}

\pagestyle{plain}
\title{JACK~---~a tool for validation of security and behaviour of Java
applications\thanks{This work is partially funded by the IST programme
of the European Commission, under the IST-2003-507894
\textsf{Inspired} and IST-2005-015905 \textsf{Mobius} projects.}}

\author{Gilles Barthe\inst{1} \and
        Lilian Burdy\and %\inst{2} \and
        Julien Charles\inst{1} \and
        Benjamin Gr\'egoire\inst{1} \and
        Marieke Huisman\inst{1} \and
        Jean-Louis Lanet\inst{2} \and
        Mariela Pavlova\inst{3}\thanks{Research done while at INRIA Sophia Antipolis.} \and
        Antoine Requet\inst{2}}
\institute{INRIA Sophia Antipolis, France \and 
           gemalto, France \and 
           Ludwig-Maximilians-Universit\"at M\"unchen, Germany}


\input {preamble}          % input user-defined commands

\begin{document}
\maketitle      
\begin{abstract}
We describe the main features of JACK (Java Applet Correctness Kit), a
tool for the validation of Java applications, annotated with JML
specifications. JACK has been especially designed to improve the
quality of trusted personal device applications. JACK is fully
integrated with the IDE Eclipse, and provides an easily accessible
user interface. In particular, it allows to inspect the generated
proof obligations in a Java syntax, and to trace them back to the
source code that gave rise to them. Further, JACK provides support for
annotation generation, and for interactive verification. The whole
platform works both for source code and for bytecode, which makes it
particularly suitable for a proof carrying code scenario.
\end{abstract}
\input{introduction}        
\input{jml}
\input{architecture}
\input{user-interface}
\input{annot-gen}
\input{bytecode}
\input{interactive-verif}
\input{conclusion}

\bibliographystyle{plain}
\bibliography{bibli,fmco,../specification,/user/mhuisman/home/Research/Everest/Biblio/everest,/user/mhuisman/home/Research/Everest/Biblio/crossrefs}

\end{document}

